---
title: "Module 3 - ANOVA"
author: "TVZ"
output:
  pdf_document:
    toc: true
    toc_depth: 3
  word_document:
    toc: true
    toc_depth: '3'
  html_document:
    df_print: paged
    toc: true
    toc_depth: 3
urlcolor: blue
editor_options:
  markdown:
    wrap: 80
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r results='hide'}
# Upload the libraries used in this document
#library("ggplot2")
#library("tinytex")
#library("extraDistr")
#library("sysid")
library("DescTools")
library('effectsize')
# library('PairedData')
library('pwrss')
library('pwr')
#library("psych")
#library('multcomp')
```

This notebook is extra! Do what you can, if you can. Do what you find
interesting. It should be useful for you as a reference while you work on your
own analysis problems, most of which will involve some kind of ANOVA or
regression.

# Learning Objectives

Analysis of Variance (ANOVA) is a statistical method used to compare means
across two or more groups to determine if there are any significant differences
between those groups. We will first discuss the ANOVA model, then the process of
*partitioning the variance*, which is the theoretical foundation for the ANOVA
procedure. Following that, we will fit an ANOVA model and determine if a
treatment has significant effects and perform diagnostics to determine if the
assumptions of the ANOVA model are satisfied. Then we discuss the concepts of
power and effect size, and how these concepts vary from the same concepts
encountered in $z$- and $t$-tests. We will talk about post-hoc and planned
contrasts, and end with the analysis of factorial and repeated measures designs.

At the end of this notebook you should be able to:

1.  Identify scenarios in which Analysis of Variance (ANOVA) is the appropriate
    statistical tool for comparing means across multiple groups.
2.  Identify the key assumptions of ANOVA (normality, homogeneity of variances,
    independence of observations) and explain the consequences of violating
    them.
3.  Compute the F-statistic and explain how its magnitude relates to the
    rejection of the null hypothesis of equality of means.
4.  Distinguish clearly between the application and interpretation of One-Way
    and Two-Way ANOVA.
5.  Define a statistical interaction effect in the context of Two-Way ANOVA and
    explain why a significant interaction dictates the primary focus of
    interpretation.
6.  Structure a dataset appropriately in `R` and use built-in `R` functions (or
    common packages like `aov` or `lm`) to execute both One-Way and Two-Way
    ANOVA.
7.  Generate and interpret necessary diagnostic plots (e.g., Q-Q plots,
    residuals vs. fitted plots) to formally check the assumptions of the ANOVA
    model.
8.  Summarize and report the key results of the ANOVA, including degrees of
    freedom, F-statistic, and p-value, in an academic or report-ready format.
9.  Explain the difference between planned (a priori) contrasts and post-hoc
    tests, and justify when to use each approach based on the research question.
10. Apply a common post-hoc procedure (such as Tukey's HSD) using an appropriate
    `R` package and interpret the resulting pairwise comparisons.
11. Formulate specific hypotheses and assign appropriate contrast coefficients
    to test targeted comparisons between means (e.g., simple, repeated, or
    polynomial contrasts).
12. Recognize the problem of inflated Type I error due to multiple comparisons
    and explain the rationale behind common correction methods (e.g.,
    Bonferroni, Holm).
13. Identify main effects and interactions from a plot of condition means.
14. Perform a factorial ANOVA for data from experimental designs with more than
    1 categorical predictor variables.
15. Perform a repeated measures ANOVA for data from experimental designs with a
    single repeated categorical predictor variable.

# Upload GSS Data

```{r}
data <- read.csv("GSS_1980.csv",header=TRUE)
data$X <- NULL
```

# Introduction to ANOVA

The hypothesis tests that we have considered up to this point have been focused
on only one or two means, with the goal of arguing against the hypothesis that
either one mean comes from a sample from a population with a stated null mean
$\mu_0$ or that the difference between two means is due to sampling variability
(chance: $\mu_1 = \mu_2$). But what about the situation where we have two or
more means?

ANOVA examines how a (numeric) dependent variable changes across levels of one
or more *categorical* independent variables. For example, consider the GSS_1980
dataset. Let's look at respondents' income as a function of their reported
health.

```{r}
# Pull out variables of interest
money <- data$REALRINC
siblings <- data$SIBS
age <- data$AGE
happy <- data$HAPPY

# Make sure the health and happiness ratings are categorical (super important)
health <- as.factor(data$HEALTH)
happy <- as.factor(data$HAPPY)

# Redefine the data object
data <- data.frame("money" = money,"health"= health,"age"=age,"siblings"=siblings,"happy"=happy)
```

Let's focus on respondents' age as a function of reported health, under the
assumption that people who report poor health (health=4) will tend to be older
than those who report good health (health \< 4).

```{r}
# Plot age as a function of health
plot(health,age)
```

This *box and whisker* chart tends to confirm our suspicions about the
relationship between age and reported health. But we would like to say that the
differences in the means across levels of reported health are *significantly*
different. To do this we will perform an analysis of variance (ANOVA).

The ANOVA null hypothesis is $$
H_0: \mu_1 = \mu_2 = \cdots = \mu_J
$$ for $J$ groups or $J$ *levels* of an independent variable. This hypothesis is
called the *omnibus null*: we're going to design a test for equality of all the
groups in our experiment. The alternative to this hypothesis is $$
H_1: \text{ At least one } \mu_j \text{ is different from the others}.
$$

When we reject $H_0$ we will not know why we rejected it: ANOVA alone will not
tell us much about what's going on in each group. We'll need to do other things
to figure that out (we'll discuss these things later).

### 1. Your turn!

Using your own dataset, select the dependent variable and a categorical
independent variable with at least 3 levels. Create a box-and-whisker plot of
the dependent variable as a function of the independent variable. State your
null and alternative hypotheses for an ANOVA test of equality of means across
the levels of your independent variable.

```{r}
# Insert your code here
```

## Can't We Just Do $t$-Tests?

Your immediate instinct (coming off of several weeks of doing nothing but
$t$-tests) might be that the way to test $$
H_0: \mu_1 = \mu_2 = \cdots = \mu_J
$$ is to contrast all pairs of means across the $J$ groups by doing
$\binom{J}{2} = J(J-1)/2$ independent samples $t$-tests. While this might not be
a bad idea, it has one serious shortcoming. What is the probability of making at
least one Type I error across all of those pairwise tests? That is, if we want
our probability of erroneously rejecting $H_0: \mu_i = \mu_j$ for one or more
pairs of means not to exceed $\alpha$, will the pairwise $t$-test strategy be
the best approach?

Assume that $J=4$. Then $\binom{4}{2} = \dfrac{4 \cdot 3}{2} = 6$. For each of
those 6 tests we'll set $\alpha = 0.05$. The probability of *not* making a Type
I error on any one test is $1 - \alpha = 0.95$. Under the assumption of
independence across the groups (which we need for the independent samples
$t$-test *and* the ANOVA), the probability of not making a single Type I error
across the 6 tests is $(1 - \alpha)^6$. Conversely, the probability of making
one or more Type I errors is $1 - (1 - \alpha)^6$.

Because $1 - \alpha$ is less than 1, raising it to higher and higher powers
results in a smaller and smaller probability of not making an error. In turn,
the probability of making an error must therefore increase for all values of
$\alpha < 1$.

```{r}
M <- 1:100 # number of tests
alpha = 0.05

# Probability of at least one error
prob.1 <- 1 - (1 - alpha)^M

plot(M,prob.1,type='p',xlab="Number of Tests",ylab="Probability of at Least One Error",cex=.25)
abline(h=c(0.05,1.0),col='purple4',lty=2)
```

This graph shows the probability of making at least one Type I error when
$\alpha=0.05$ as a function of the number of pairwise tests to be performed.
This probability grows quickly with the number of tests to be performed. By the
time the number of tests reaches 14 the probability is greater than half. By the
time it reaches 59 the probability is greater than 95%. Fourteen tests
corresponds to around $J=6$ groups; 59 tests corresponds to $J=12$ groups (more
than you should be considering in a single experimental design). Bottom line:
pairwise $t$-testing is not a good solution for testing equality of more than
two means.

This $t$-test strategy (pairwise $t$-testing) is something we will encounter
later in the form of post-hoc testing. Here, however, we will consider an
alternative strategy (ANOVA) that "rewrites" the null hypothesis $H_0$ as a
question about variances rather than means. Remember that the F-distribution
describes the random variable that arises from a ratio of two variances and that
the F-test is used to examine the hypothesis $H_0: \sigma^2_1 = \sigma^2_2$, or
equivalently $H_0: \sigma^2_1/\sigma^2_2 = 1$.

## The Logic of ANOVA

Consider the distribution of age measurements across the four health levels.

```{r Anova logic}
# Plot age as a function of reported health
plot(age~as.numeric(health),xlab="Reported Health",ylab="Age",cex=.5)

# Add the means of each group to the plot
means <- aggregate(age,by=list(health),mean)
points(means$Group.1,means$x,pch='x',col='blue',cex=1.5)
lines(means$Group.1,means$x,col='blue')
```

We are interested in whether the mean ages at each level of reported health are
different. The blue Xs mark the group means that are the focus of our attention.
Remember that (almost) every statistical procedure is based on a *model*, an
assumption about how the data were generated. (Some models are more interesting
than others.) You need to understand the ANOVA model before we proceed.

# The ANOVA Model

Write $X_{ij}$ to indicate the age of respondent $i$ in group $j$. The ANOVA
model states that every observation $X_{ij}$ within group $j$ should be
approximately equal to the mean $\mu_j$ of the population from which that group
was sampled, plus some amount of noise $e_{ij}$. So $X_{ij} = \mu_j + e_{ij}$,
where $e_{ij} \sim \mathcal{N}(0,\sigma^2_e)$. We assume that the $e_{ij}$s are
independent and identically distributed with common variance $\sigma^2_e$. This
means the $X_{ij}$s must also be independent and normally distributed with mean
$\mu_j$ and variance $\sigma^2_e$. We call $\sigma^2_e$ the error variance: the
only reason that $X_{ij}$ doesn't exactly equal $\mu_j$ is because of sampling
error. An important thing to recognize is that error is inexplicable. We don't
know where it comes from, only that it makes things messy.

Notice the homoskedasticity assumption: the error variance $\sigma^2_e$ is the
same for each group. If the null hypothesis
$H_0: \mu_1 = \mu_2 = \cdots = \mu_J$ is true and all the means are the same,
and if the samples all come from populations with variance $\sigma^2_e$, then
the samples have all been drawn from the same normal distribution with mean
$\mu = \mu_j,~j=1,\ldots,J$, called the *grand mean*, and variance $\sigma^2_e$.

For reasons that will become clear later, we can rewrite the model by adding
zero as $$
X_{ij} = \mu + (\mu_j - \mu) + e_{ij} = \mu + \tau_j + e_{ij},
$$ where $\tau_j = \mu_j - \mu$ is called the *effect* of being in group $j$
(sometimes called the effect of treatment level $j$). If $H_0$ is true, all the
effects are zero, because $\mu=\mu_j$ for all $j$.

These are the most important assumptions about the ANOVA model:

1.  All groups are independent.
2.  Individuals are randomly assigned to treatment conditions.
3.  The measurements $X_{ij}$ are normally distributed.
4.  The error variance $\sigma^2_e$ is the same over all groups
    (homoskedasticity).

The ANOVA model predicts that each individual's measurement should be equal to
the population mean for the group to which the individual was assigned. We don't
know what $\mu_j$ is so we will estimate it with the group mean
$\overline{X}_j$. This means that the predicted measurement for each individual
in group $j$ is $\overline{X}_j$.

Go back to the figure in the code chunk `ANOVA logic`. The ANOVA model allows us
to reformulate the null hypothesis of equality of means in terms of predicted
and error variance. If $H_0$ is true, the predicted variance, the variance of
all the predicted scores, should be approximately the same as the error or
unexplained variance. To see how this is true, we need to *partition* the total
variance in the measured variable.

## Partitioning Variance

Consider the age variable in the GSS dataset. To begin, we will look at the
variance of age without reference to the reported health of any of the
respondents. This is the *total variance*:

```{r}
# Total variance of age
var(data$age)
```

Remember that the variance of a set of measurements is the sum of squared
deviations of the measurements from the group mean divided by the degrees of
freedom, or $$
s^2 = \dfrac{\sum_{i=1}^N (X_i - \overline{X})^2}{N-1}.
$$

For the rest of this discussion we will focus on the sum of squares
$SSX = \sum_{i=1}^N (X_i - \overline{X})^2$, the numerator of the variance of
$X$. When we look at the sum of squares of a variable over all observations of
that variable without regard to the experimental design or who is assigned to
which group we call it the *total* sum of squares. Call $N_{\text{Total}}$ the
total number of respondents and $SS_{\text{Total}}$ the total sum of squares.

```{r}
N.Total <- length(age) # all the ages
SS.Total <- var(age)*(N.Total - 1) # the numerator of the total variance
```

Now we need to modify our notation a bit to take into account the fact that
associated with each respondent's age there is also a reported health. Use the
subscript index $i$ to indicate a particular respondent and a second subscript
index $j$ to refer to the health "group" they are in, so the age of respondent
$i$ with reported health $j$ is $X_{ij}$. Each group $j=1,2,\ldots,J$ has $N_j$
respondents in it, and the number of groups is $J$ (which equals 4 in this case,
but we'll just write $J$ to keep things as general as possible). So the
respondents in group $j$ are $i = 1, 2, \ldots, N_j$.

Within each group $j$ we can compute the group mean $\overline{X}_j$ and group
variance $s^2_j$ in the usual way: $$
\overline{X}_j = \dfrac{\sum_{i=1}^{N_j} X_{ij}}{N_j},
$$ and $$
s^2_j = \dfrac{\sum_{i=1}^{N_j} (X_{ij} - \overline{X}_j)^2}{N_j - 1}.
$$

```{r}
# Compute the group statistics and store them in vectors
group.means <- aggregate(age,by=list(health),mean)
group.vars <- aggregate(age,by=list(health),var)

# Examine the group characteristics
Desc(age~health)
```

(The `Desc()` function from the `DescTools` package provides a nice summary of
the group statistics, including the group means and variances, the group sizes,
a box-and-whisker plot of age as a function of health, and a plot of the group
means. It also gives the results of a Kruskal-Wallis test, which is a
nonparametric alternative to ANOVA. You don't know yet what a Kruskal-Wallis
test is, but you can probably guess from the $p$-value it returns that there are
significant differences among the group means.)

We need to also compute the "grand mean" $\overline{\overline{X}}$, the mean of
all the ages across all the groups. If all the group sizes $N_j$ are equal we
can express the grand mean as the mean of the means, or $$
\overline{\overline{X}} = \dfrac{\sum_{j=1}^J \overline{X_j}}{k}.
$$ However, if the group sizes are not equal then we have to weight each group
mean by the group size. To do this you can work backward from the group means
using the formula for a [weighted
mean](https://en.wikipedia.org/wiki/Weighted_arithmetic_mean) or just compute
the mean over all the individual observations. Our notation for this requires an
additional summation sign: $$
\overline{\overline{X}} = \dfrac{\sum_{j=1}^J \sum_{i=1}^{N_j} X_{ij}}{N_{\text{Total}}}.
$$ It shouldn't be hard to figure out this notational twist: set $j=1$ for the
first group and add up all the values in the group ($X_{i,1}$). Then increment
$j=2$ and add up all the values in the second group and add that to the sum of
the first group. Continue incrementing $j$ until you've added up all the values
of the variable.

```{r}
# Grand mean when groups are equal or unequal
grand.mean <- mean(age)
```

Now rewrite the total sums of squares $SS_{\text{Total}}$ using this new
notation by adding up the squared deviations within the first group, then the
second, and so on. Then $SS_{\text{Total}}$ becomes $$
SS_{\text{Total}} = \sum_{j=1}^J \sum_{i=1}^{N_j} (X_{ij} - \overline{\overline{X}})^2.
$$

Let's go back to the figure and focus in on one respondent's measurement in
particular, which I've colored in orange. I've also added the grand mean to the
plot (the horizontal dashed line).

```{r}
# Plot age as a function of reported health
plot(age~as.numeric(health),xlab="Reported Health",ylab="Age",cex=.5)

# Add the means of each group to the plot
points(group.means$Group.1,group.means$x,pch='x',col='blue',cex=1.5)
lines(group.means$Group.1,group.means$x,col='blue')
abline(h=mean(age),lty=2,lwd=2) # grand mean

# Highlight observation 619
unsub <- 619
points(health[unsub],age[unsub],pch=19,col='orange',cex=1)
```

ANOVA decomposes total variance into: 1. **Between-group variance** (explained
by group differences) 2. **Within-group variance** (residual or error)

The total deviation of respondent 619's age $X_{619,4}$ from the grand mean
$\overline{\overline{X}}$ can be broken down into two components: the difference
between Group 4's mean $\overline{X}_4$ from the grand mean
$\overline{\overline{X}}$ -- the difference between the blue X and the
horizontal dashed line, between-group deviation -- and the difference between
the individual's age $X_{619,4}$ from the group mean $\overline{X}_4$ -- the
difference between the orange dot and the blue X, within-group deviation.
Mathematically, this is expressed as $$
X_{ij} - \overline{\overline{X}} = (\overline{X}_j - \overline{\overline{X}}) + (X_{ij} - \overline{X}_j).$$

Squaring both sides and summing over all respondents and all groups gives
\begin{eqnarray*}
\sum_{j=1}^J \sum_{i=1}^{N_j} (X_{ij} - \overline{\overline{X}})^2 & = & \sum_{j=1}^J \sum_{i=1}^{N_j} (\overline{X}_j - \overline{\overline{X}})^2 + \sum_{j=1}^J \sum_{i=1}^{N_j} 2 (\overline{X}_j - \overline{\overline{X}})(X_{ij} - \overline{X}_j) + \sum_{j=1}^J \sum_{i=1}^{N_j} (X_{ij} - \overline{X}_j)^2 \\
& = & \sum_{j=1}^J \sum_{i=1}^{N_j} (\overline{X}_j - \overline{\overline{X}})^2 + 2 \sum_{j=1}^J (\overline{X}_j - \overline{\overline{X}}) \sum_{i=1}^{N_j} (X_{ij} - \overline{X}_j) + \sum_{j=1}^J \sum_{i=1}^{N_j} (X_{ij} - \overline{X}_j)^2 \\
& = & \sum_{j=1}^J \sum_{i=1}^{N_j} (\overline{X}_j - \overline{\overline{X}})^2 + \sum_{j=1}^J \sum_{i=1}^{N_j} (X_{ij} - \overline{X}_j)^2.
\end{eqnarray*}

The term in the middle that is multiplied by 2 that we get after expanding the
product on the right side is zero because the sum of deviations from the group
mean $\sum_{i=1}^{N_j} \left( X_{ij} - \overline{X}_j \right)$ is always zero.

The left side is the total sum of squares $SS_{\text{Total}}$. The first term on
the right side is called the *between-groups* sum of squares
$SS_{\text{Between}}$, which measures how much of the total variance is
explained by differences between group means. The second term on the right side
is called the *within-groups* sum of squares $SS_{\text{Within}}$, which
measures the variance within each group that is not explained by group
membership (the error variance). So we have
$$SS_{\text{Total}} = SS_{\text{Between}} + SS_{\text{Within}}.$$ If we divide
each term by their associated degrees of freedom, we say that the total variance
is partitioned into variance explained by group differences and unexplained
variance (error).

## Does this really work?

Compute the three sums of squares for the age variable.

```{r}
# Determine the numbers of observations and groups
N.total <- length(age) # all the ages
N.groups <- aggregate(age,by=list(health),length)$x
J <- length(N.groups) # number of groups

# Compute the means
grand.mean <- mean(age)
group.means <- aggregate(age,by=list(health),mean)$x

# Total sum of squares
SS.total <- sum((age - grand.mean)^2)

# Between-groups sum of squares
SS.between <- sum(N.groups * (group.means - grand.mean)^2)

# Within-groups sum of squares
SS.within <- sum(aggregate(age,by=list(health),
             function(x) sum((x - mean(x))^2))$x)
# Alternative:
# SS.within <- sum((N.groups - 1)*     
#              aggregate(age,by=list(health),var)$x )

# Check that the sums of squares add up
cat('SS.total = ',SS.total)
cat('\nSS.between = ',SS.between)
cat('\nSS.within = ',SS.within)
cat('\nSS.between + SS.within = ',SS.between + SS.within)
```

## 2. Your turn!

Using your own dataset, compute the total, between-groups, and within-groups
sums of squares for your dependent variable. Verify that the between and within
sums of squares add up to the total.

```{r}
# Insert your code here
```

## The ANOVA F-Test

Remember that $\overline{X}_j$ is the predicted value for each individual in
group $j$. For the between-groups variance (sometimes called predicted
variance), we will replace all the individual measurements $X_{ij}$ with their
predicted values $\overline{X}_j$. For the within-groups variance (sometimes
called residual or error variance), we will focus on the extent to which the
predicted values $\overline{X}_j$ differ from the actual measurements $X_{ij}$.

*If the null hypothesis is true*, the group means should be approximately equal
or $\overline{\overline{X}} = \overline{X}_j$ for all $j$, so the only reason
that any person's score should differ from the grand mean
$\overline{\overline{X}}$ is because of sampling variance or error. Therefore,
the between-groups variance should estimate the population variance
$\sigma^2_e$. Keep in mind that, under the assumption of homoskedasticity, the
sample variances within each group are also estimates of $\sigma^2_e$, so the
within-groups variance should also estimate $\sigma^2_e$. Thus, if $H_0$ is
true, the between-groups variance and the within-groups variance should be
approximately equal.

*If the null hypothesis is false*, at least one group mean $\overline{X}_j$ is
different from the others, so the between-groups variance should be large
relative to the within-groups variance. This leads us to the key idea behind
ANOVA: we can test the null hypothesis of equality of means by comparing the
between-groups variance to the within-groups variance. If the between-groups
variance is significantly larger than the within-groups variance, we will reject
$H_0$. If the two variances are approximately equal, we will fail to reject
$H_0$. See the section below, "Back to the ANOVA Model," for more about this.

To perform this comparison we need to convert the sums of squares into variances
by dividing each sum of squares by its associated degrees of freedom. The
between-groups variance has $J-1$ degrees of freedom (one less than the number
of groups). This is because we are replacing each $X_{ij}$ with its group mean
$\overline{X}_j$, and so the only degrees of freedom we have are selecting the
values of the group means: the mean of these means must equal the grand mean
$\overline{\overline{X}}$ and so the value of the last group mean is determined
by the values of the other $J-1$ group means.

The within-groups variance has $N_{\text{Total}} - J$ degrees of freedom (the
total number of observations minus the number of groups). Within each group we
have $N_j - 1$ degrees of freedom as for the usual computation of a sample
variance. The total degrees of freedom for the within-groups variance is
therefore the sum of the degrees of freedom across all groups, or $$
\sum_{j=1}^J (N_j - 1) = \left(\sum_{j=1}^J N_j\right) - J = N_{\text{Total}} - J.
$$

Dividing each sum of squares by its degrees of freedom gives us the *mean
square* for each source of variance:
$$MS_{\text{Between}} = \dfrac{SS_{\text{Between}}}{J-1},$$ and
$$MS_{\text{Within}} = \dfrac{SS_{\text{Within}}}{N_{\text{Total}} - J}.$$ These
MS values give two different estimates of the variance $\sigma^2_e$ *if* $H_0$
is true. The null hypothesis $$
H_0: \mu_1 = \mu_2 = \cdots = \mu_J
$$ can therefore be evaluated by the equivalent null hypothesis $$
H_0: \sigma^2_{\text{Between}} = \sigma^2_{\text{Within}}
$$ or $$
H_0: \dfrac{\sigma^2_{\text{Between}}}{\sigma^2_{\text{Within}}} = 1.
$$

Estimating the variances with the mean squares leads to the ANOVA $F$-statistic,
which is the ratio of the mean squares: $$
F(J-1,N_{\text{Total}} - J) = \dfrac{MS_{\text{Between}}}{MS_{\text{Within}}},
$$ with $J-1$ numerator and $N_{\text{Total}} - J$ denominator degrees of
freedom. If $H_0$ is true, this ratio should be approximately 1. If $H_0$ is
false, the ratio should be significantly larger than 1.

# Fit an ANOVA Model

Going back to the GSS data, we can fit an ANOVA model for age as a function of
reported health using the `aov()` function in R. The commands in this chunk fit
the model (but we won't see that yet) and produces a table of group means, group
sizes, and the grand mean.

```{r}
# Fit a one-way ANOVA model
age.model <- aov(age~health,data=data)
model.tables(age.model, type = "means")
```

## The ANOVA Summary Table

A lovely way to present the results of an ANOVA is via the ANOVA summary table,
which can be obtained using the `summary()` function in `R`.

```{r}
summary(age.model)
```

Summary tables are no longer as popular as they used to be, because they take up
a lot of space and don't convey much information beyond the $F$-statistic and
the associated p-value. However, they do provide a nice way to see the
partitioning of variance that is the foundation of ANOVA and to confirm that the
computations that you think `R` is doing are correct.

In the summary table above, you can see the partitioning of variance into
between-groups and within-groups components. The between-groups component is
identified by the name of the independent variable (health), and the
within-groups component is identified by the label "Residuals." (Residual sum of
squares is another name for within-groups sum of squares; you will also see it
referred to as error sum of squares.)

You can verify for yourself that these sums of squares are the same as the ones
we computed earlier and that the Residual and health sums of squares add up to
the total sum of squares. Another thing to notice is that the degrees of freedom
for each source of variance sum to the total degrees of freedom,
$N_{\text{Total}} - 1 = (J-1) + (N_{\text{Total}} - J)$.

The $F$-statistic is the same that we would use to test the null hypothesis
$H_0: \sigma_1^2 = \sigma_2^2$, so $F = s_1^2/s_2^2$, the ratio of two sample
variances. In ANOVA, the sample variances that estimate the population variances
$\sigma_{\text{Between}}$ and $\sigma_{\text{Within}}$ are $MS_{\text{Between}}$
and $MS_{\text{Within}}$, which should be approximately equal if
$H_0: \mu_1 = \mu_2 = \cdots = \mu_J$ is true. Thus, the $F$-statistic is
$MS_{\text{Between}}/MS_{\text{Within}}$. This statistic follows an
$F$-distribution with $J-1$ numerator and $N_{\text{Total}} - J$ denominator
degrees of freedom.

The degrees of freedom (Df) for the numerator and denominator of the
$F$-statistic and p-value are presented in the summary table. The probability of
obtaining an $F$-statistic as large or larger than the observed value under the
null hypothesis is very small ($p$-value \< $\alpha$, if $\alpha = 0.05$), so we
reject the null hypothesis of equality of means across reported health levels
and conclude that at least one group mean is different from the others.

## 3. Your turn!

Using your own dataset, fit an ANOVA model for your dependent variable as a
function of your categorical independent variable. Present the ANOVA summary
table and interpret the results.

```{r}
# Insert your code here
```

## Back to the ANOVA Model

Remember that the ANOVA model predicting $X$ from the group means can be written
as $$
X_{ij} = \mu + \tau_j + e_{ij},
$$ where we defined $\mu$ to be the grand mean, $\tau_j$ is the effect of being
in group $j$, and $e_{ij}$ is the error term which we assume is normal with mean
0 and variance $\sigma_e^2$. We defined the effect $\tau_j$ to be
$\tau_j = \mu_j - \mu$, the difference between the group mean and the grand
mean. If $H_0$ is true, all the effects $\tau_j = 0$; "no effect of health."

### The Sampling Distributions of the Mean Squares

Because $MS_{\text{Within}}$ and $MS_{\text{Between}}$ are sample
statistics they have their own sampling distributions (which are
related to $\chi^2$ distributions). The means of these distributions are given
by $$
E[MS_{\text{Within}}] = \sigma^2_e,
$$ and $$
E[MS_{\text{Between}}] = \sigma^2_e + \dfrac{\sum_{j=1}^J N_j \tau_j^2}{J-1}.
$$

If $H_0: \mu_1 = \cdots = \mu_J$ is true, all the $\tau_j$s equal 0, so we would
expect that $$
F = \dfrac{MS_{\text{Between}}}{MS_{\text{Within}}} \approx \dfrac{\sigma^2_e}{\sigma^2_e} = 1.
$$
If $H_0$ is false, at least one $\tau_j \neq 0$, so we would expect that the $F$ statistic will be significantly greater than one, because the squared effects, the squared $\tau_j$s, will make the numerator larger. 

### Interpreting the Model Parameters

The interpretation of the model parameters $\mu$ and $\tau_j$ depends on how the
independent variable is coded. In the formulation above we used *effect coding*,
where $\mu$ is the grand mean and the effects are expressed as differences from
the grand mean.

An alternative is to define $\mu$ to be the mean of a reference or control
group, say group 1, and define the effects as differences from that group. So we
would have $$
X_{ij} = \mu_1 + (\mu_j - \mu_1) + e_{ij} = \mu_1 + \delta_j + e_{ij},
$$ where again $\delta_j$ is the effect of being in group $j$, but now defined
as the difference between the group mean and the mean of the reference group. If
$H_0$ is true, all the effects $\delta_j = 0$; "no effect of health."

I bring this up to emphasize the fact that ANOVA is a kind of regression model,
just like the simple regressions you have done in earlier classes where your
goal was to predict a variable $Y$ from a single continuous variable $X$. The
only difference is that here we are predicting $Y$ from a categorical variable
with $J$ levels.

The `aov()` function in `R` fits the ANOVA model using *dummy coding*, where the
first level of the independent variable is treated as the reference group. You
can see this by using the `summary.lm()` function on the fitted ANOVA model.

```{r}
summary.lm(age.model)
```

The intercept is the mean of the health=1 (excellent) group and the coefficients
for the other levels of health are the differences between those group means and
the reference group mean. For example, the coefficient for health=2 (good
health) is 0.9109, which means that the mean age for respondents reporting good
health is 0.9109 years older than the mean age for respondents reporting
excellent health.

Associated with each coefficient is a $t$-statistic and $p$-value testing the
null hypothesis that the coefficient equals zero. You should interpret these
tests with caution, however, because multiple comparisons increase the
probability of Type I error. We will discuss this issue more when we get to post
hoc tests.

## 4. Your turn!

Using your own dataset, use the `summary.lm()` function to examine the
coefficients of your fitted ANOVA model. Interpret the coefficients and their
associated t-statistics and p-values.

```{r}
# Insert your code here
```

# ANOVA Diagnostics

ANOVA requires that the data meet certain assumptions, including normality of
the residuals and homoskedasticity (equal variances across groups). Also, the
presence of outliers may produce distorted results: outliers will increase the
$SS_{\text{Within}}$, reducing the $F$-ratio, and consequently reducing the
effect size and power. Now that we've fit the model we can look for signs that
these assumptions hold.

## Outliers

We need to first check to see if we have outliers and, if so, whether they will
influence the results of the analysis. You can do this by eye, looking at the
box-and-whisker plots of the dependent variable (age) as a function of the
independent variable (health). You can also apply some of the procedures for
identifying outliers in a sample to the observations within each group (e.g.,
you might label observations $X_{ij}$ more than $3 s_j$ from the group mean
$\overline{X}_j$, or observations in the lowest or highest quantiles of the data
distribution as outliers).

This chunk of code looks for outliers defined as more than 3 standard deviations
$s_j$s from the group mean and creates new variables `age.filtered` and
`health.filtered` from which those outliers have been removed. (Note that there
are [lots of `R` resources](https://statsandr.com/blog/outliers-detection-in-r/)
for outlier detection)

```{r}
# Compute the group means and standard deviations
# Group means were computed above
# group.means <- aggregate(age,by=list(health),mean)
sds <- aggregate(age,by=list(health),sd)$x

# There are J groups (J is determined above)
# J <- length(unique(health))

# Evaluate each group for outliers that are 3 sds from the group
# means
# Store the outliers in a list
outliers <- list()
age.filtered <- health.filtered <- vector()
for (i in 1:J) { 
  # Compute upper and lower bounds as group.mean +/- 3 group.sd
  group.data <- age[health == i]
  group.mean <- group.means[i]
  group.sd <- sds[i]
  lower.bound <- group.mean - 3*group.sd
  upper.bound <- group.mean + 3*group.sd

  # Which values are farther than 3 group.sd from the group.mean  
  group.outliers <- group.data[group.data < lower.bound | 
                              group.data > upper.bound]
  keep.data <- group.data[group.data >= lower.bound & 
                              group.data <= upper.bound]
  
  # How many outliers in the group?
  n.outlier <- length(group.outliers)
  
  # If none, set the values to -1
  if(n.outlier==0) group.outliers<--1
  
  # Store the results in the list
  outliers[[i]] <- c("group"=i,"N"=n.outlier,"values" = group.outliers)
  
  # Remove the outliers to make a new dataset
  age.filtered <- c(age.filtered,keep.data)
  health.filtered <- c(health.filtered,
                      rep(i,times=length(keep.data)))
}
outliers
```

This method identifies 5 outliers: 3 in the health=1 group, 1 each in the
health=2 and 3 groups, and none in the health=4 group. One reasonable approach
is to do a *sensitivity analysis*: redo the analysis after removing the
outliers. If nothing changes then you don't need to worry (much) about them. If
your conclusions change then you have a problem. As an expert in your field, you
will need to use some judgment about how to handle these observations. If you
intend to remove them from the dataset, keep in mind that you will need to have
determined in advance what the criteria for exclusion are (such as $\pm 3 s_j$)
and preregister these criteria.

```{r}
# Do the ANOVA again with the filtered data
health.filtered <- as.factor(health.filtered)
filtered.model <- aov(age.filtered~health.filtered)
summary(filtered.model)
```

The results of the ANOVA after removing the outliers are very similar to the
original results, so we can be reasonably confident that the outliers are not
having a big influence on the results.

### Cook's Distance

An outlier can be characterized by its leverage and its influence. Leverage
quantifies the influence that a predictor/independent variable (health) has on
the dependent variable (age). It is a function only of the predictor variable
and is independent of the dependent variable. Influence measures the impact that
a single observation has on the estimated model parameters. Observations with
high influence, which is a function of both leverage and the distance of an
observation from its group mean, can drastically change the outcome of an
analysis and are therefore the most problematic. Influence is commonly measured
with *Cook's distance.*

An observation is considered influential if its Cook's distance $D_{ij}$ is
greater than $4/(N_{\text{Total}}-J-1)$. While the formula for $D_{ij}$ is
beyond the scope of this class, you won't be surprised to learn that there are
`R` functions that will compute it for you.

```{r}
# Compute Cook's distance for each observation and flag those that have Cook's distances greater than 4/(N-k-1).
D.limit <- 4/(N.Total - J - 1)
cooks.d <- cooks.distance(age.model)
influential <- which(cooks.d > D.limit)

# Flag the influential observations in each group
influence <- data.frame("health"=health[influential],  
                                 "age"=age[influential],
                                 "cooks.d"=cooks.d[influential])

# List the influential observations by group
inf.groups <- list()
  for (i in 1:J) 
      inf.groups[[i]] <-    
      c("group"=i,"values"=influence[influence[,1]==i,2])
inf.groups
```

You can see that the Cook's distance measure has identified more observations
that are highly influential, and those that were identified by distance from the
mean are included among these. Also while outliers identified by distance are
all older respondents the Cook's distance measure also identifies younger
respondents.

In summary, there are potentially troublesome outliers in the GSS data. How we
deal with them is an open question. If, after checking for coding errors or
unique circumstances that led to the measurements, nothing seems amiss, then you
can perform a sensitivity analysis as above. If the results are different with
vs. without the influential points you have a fairly serious problem. *You must
report any conflict between the results obtained with vs. without influential
points.*

### 5. Your turn!

Conduct an outlier analysis for your data. If you determine that there are no
outliers, describe this finding. If outliers are present, perform a sensitivity
analysis. Describe your results. If the sensitivity analysis indicates a
problem, discuss what you might do to resolve it.

```{r}
# Insert your code here
```

## Normality and Homoskedasticity

The ANOVA model assumes that errors are normally distributed. Error is estimated
as the difference between the observed values $X_{ij}$ and the predicted values
$\hat{X}_{ij} = \overline{X}_j$ from the model. Because $$
X_{ij} = \hat{X}_{ij} + e_{ij},  
$$ the *residual* $X_{ij} - \hat{X}_{ij} = e_{ij}$. It is the residuals (errors)
that we examine for normality and homoskedasticity, because the model says that $e_{ij} \sim \mathcal{N}(0,\sigma^2_e)$.

A residual plot looks at the residuals $e_{ij}$ as a function of the predicted
values $\hat{X}_{ij}$. These variables are returned as part of the `aov` model
object.

```{r}
# Residual plot
plot(age.model$fitted.values,age.model$residuals,
     xlab="Fitted Values",ylab="Residuals",cex=.5)
abline(h=0)
```

Notice that there are only 4 fitted (or predicted) values, because (again) the ANOVA
model says that an individual's measurement $X_{ij}$ is predicted by the mean of
the group $\overline{X}_j$ to which that individual belongs:
$\hat{X}_{ij} = \overline{X}_j$. The horizontal line is placed at a residual
value of 0.

Things to look for in a residual plot are:

1.  Symmetry about 0. Errors are normally distributed with mean 0. If the
    residuals are asymmetric around 0, there is a failure of the normality
    assumption. If the residuals for each group are not centered at 0 then there
    is structure in the data that has not been captured by the ANOVA model.
2.  Approximately equal ranges in residuals across groups. If the ranges are
    evidently different, variance within each group differs across the groups
    and homoskedasticity is violated.
3.  Outliers. Outliers and influential values in the raw data will appear as
    outliers in the residuals because the model will have difficulty predicting
    them well.

These features of a residual plot are indirect evidence of violations of
assumptions. We can be more precise with QQ plots and statistical tests of
homoskedasticity.

### QQ plots

Remember that a QQ plot plots theoretical quantiles from a specified
distribution against the quantiles of the data. If a normal distribution is a
good fit to the data then the QQ plot will show a nice straight line through the
plotted points.

```{r}
# QQ plot of residuals
qqnorm(age.model$residuals)
# Add a best-fitting line
qqline(age.model$residuals)
```

The plot shows a good fit to the normal distribution everywhere except in the
lower quantiles of residuals. We can get even more precise by doing a
*Shapiro-Wilk test* of normality. This test focuses specifically on the deviations from the
line shown in the QQ plot.

```{r}
shapiro.test(age.model$residuals)
```

The $p$-value of the Shapiro-Wilk test statistic ($W$) is less that
$\alpha=0.05$, so we reject the null hypothesis that the residuals are normally
distributed. However, the Shapiro-Wilk test is very sensitive for large sample
sizes like the GSS dataset, so the deviation from normality, while significant,
may not be interesting.

### Levene's Test of Equality of Variances

It is not clear from the residual plot if the homoskedasticity assumption is
violated. The ranges are different across the groups but it can't be determined
just by looking if these differences are due to sampling error. To investigate
this more closely, you can perform a Levene's test. This test is robust to
violations of the normality assumption, which is something that we now need to
be sensitive to.

```{r}
# Levene's test for homogeneity of variance (in car library)
library(car)
leveneTest(age.model)
```

The $p$-value for the Levene's test is greater than $\alpha=0.05$, suggesting
that we can't reject the null hypothesis of homoskedasticity. So this assumption
appears to hold for our data.

### 6. Your turn!

Perform a residual analysis of the ANOVA model fit to your data. Construct the
residual and QQ plots and analyze them "by eye." Using Shapiro-Wilks and
Levene's tests, determine the extent to which the ANOVA assumptions hold in your
data.

```{r}
# Insert your code here
```

# Power and Effect Size

Power analysis and effect size calculations are important for understanding the
practical significance of ANOVA results. In contrast to comparisons between two
means, effect size measures for ANOVA often focus on the proportion of variance
explained by the independent variable.

## Effect Sizes

Like for the $t$-test, when the null hypothesis is false the $F$-ratio is not
distributed as an $F$. Instead, again like the $t$ statistic, it follows a
non-central $F$ distribution. So the issue of effect sizes is often cast in
terms of a noncentrality parameter. If the central $F$ has parameters $\nu$ and
$\delta$, the numerator and denominator degrees of freedom, the non-central $F$
has parameters $\nu$, $\delta$, and $\lambda$, where $\lambda$ is the
noncentrality parameter. We'll return to the computation of $\lambda$ in a
moment.

There are three common effect size measures for ANOVA: eta-squared ($\eta^2$),
omega-squared ($\omega^2$) and Cohen's $f^2$. In contrast to the $d$ (distance)
measures of effect size we used for t-tests, these are called $r$ (correlation)
measures.

Eta-squared is calculated as
$$\eta^2 = \dfrac{SS_{\text{Between}}}{SS_{\text{Total}}}.$$ It represents the
proportion of total variance in the dependent variable that is explained by the
ANOVA model. (This is the same as the $r^2$ measure used in simple regression.)

While simple, unfortunately $\eta^2$ is a biased estimate of the effect size: it
tends to overestimate the proportion of variance explained, especially for small
sample sizes. Omega-squared is a less biased estimate of effect size and is
calculated as
$$\omega^2 = \dfrac{SS_{\text{Between}} - (J - 1)MS_{\text{Within}}}{SS_{\text{Total}} + MS_{\text{Within}}}.$$

Cohen's $f^2$ is a transformation of $\eta^2$: $$
f^2 = \dfrac{\eta^2}{1 - \eta^2},
$$ and when we talk about an effect size in ANOVA we usual mean Cohen's $f$,
which is just $$
f = \sqrt{\dfrac{\eta^2}{1 - \eta^2}}.
$$

The noncentrality parameter $\lambda$ for the non-central $F$ distribution is a
function of Cohen's $f^2$: $$
\lambda = f^2 \sum_{i=1}^k N_j.
$$

These are a lot of little formulas, and I present them so you know how to use
them in `R` and to compute power using different functions in `R`.

Like the Cohen's $d$ we encountered in tests of two means, there are guidelines
for what we consider to be small, medium, or large effect sizes for $\eta^2$,
$\omega^2$ and $f^2$:

| Size           | $\eta^2,\omega^2$   |    | f    |
|:---------------|:-------------------:|--- |:----:|
|----------------|---------------------|    |------|
| Small effect   | 0.01                |    | 0.10 |
| Medium effect  | 0.06                |    | 0.25 |
| Large effect   | 0.14                |    | 0.40 |
|----------------|---------------------|    |------|

These effects are easy to compute using commands from `R`'s `effectsize`
package. Here is some code to compute $\eta^2$, $\omega^2$ and Cohen's $f$.

```{r}
# Calculate effect size
eta.2 <- eta_squared(age.model)$Eta2
omega.2 <- omega_squared(age.model)$Omega2
cohens.f <- cohens_f(age.model)$Cohens_f
```

The `effectsize` library has a bunch of random functions that you might find
useful. We can convert $\eta^2$ to $f^2$ using the `eta2_to_f2` function.

```{r}
f.2 <- eta2_to_f2(eta.2)
f <- eta2_to_f(eta.2) # A fancy sqrt function

# Check:
cat("eta^2 = ",eta.2,", omega^2 = ",omega.2," Cohen's f = ",cohens.f)
cat("\nf^2 = ",f.2," f = ",f) # returned by functions
cat("\nf^2 = ",eta.2/(1-eta.2)," f = ",sqrt(f.2)) # check by hand
```

Both $\eta^2$ and $\omega^2$ indicate that reported health explains a small
proportion of the variance in age. Cohen's $f$ also suggests a small effect
size.

### 7. Your turn!

Using your own dataset, compute the effect sizes $\eta^2$, $\omega^2$ and
Cohen's $f$ for your fitted ANOVA model. Interpret the results.

```{r}
# Insert your code here
```

## Power Analysis

We've been using the `pwrss` package for power analysis with $t$-tests. It also
provides functions for power analysis with ANOVA. For diversity, and because the
anova power computations are a bit easier, we'll use the `pwr` package for power
analysis with ANOVAs.

To compute sample size for a given level of power we need the desired effect
size, significance level, and the information necessary to compute degrees of
freedom ($J-1$).

```{r power-analysis}
# Power analysis for one-way ANOVA of age~health data
# Desired power
one.minus.beta <- 0.80

# (Desired) effect size (Cohen's f)
cohens.f <- 0.25 # medium

pwr.anova.test(k = 4, f = cohens.f, sig.level = 0.05,
               power = one.minus.beta)
```

This output indicates that to detect a medium effect size ($f = 0.25$) with 80%
power at $\alpha = 0.05$ with 4 groups, we need a total sample size of 180
respondents (45 per group $\times$ 4).

The `pwrss` package can perform a similar computation with the command
`power.f.ancova`, but rather than Cohen's $f$ it requires $\eta^2$. An ANCOVA is
a generalization of an ANOVA model in which predictor variables (like health)
can be both categorical and numeric. Here is code equivalent to that for
`pwr.anova.test` using `power.f.ancova`.

```{r}
# Convert f=0.25 to eta^2
new.eta2 <- f2_to_eta2(0.25^2)
# new.eta2 <- 0.25^2/(1 + 0.25^2) # by hand

# Compute sample size for ANOVA with 4 groups and desired power 0.80
power.f.ancova(new.eta2,factor.levels=4,power=.8,alpha=.05)
```

The `power.f.ancova` function returns a necessary total sample size of 180. It
is assumed (and desirable) that the 4 group sizes be as equal as possible: 180/4
= 45, exactly the number returned by the `pwr.anova.test` function.

### 8. Your turn!

Compute the post-hoc power of the ANOVA you performed on your own data. Compute
Cohen's $f$ directly or use a combination of the `eta_squared` and `eta2_to_f`
functions from the `effectsize` package. Then use the `pwr.anova.test` function
to compute the power of your ANOVA given your sample size, number of groups, and
effect size.

```{r}
# Insert your code here
```

# Mean Contrasts

Go back to the issue of determining *why* an omnibus $F$-test is significant:
which mean or means are different from the rest? There are two ways to handle
this. One way is reflective of the "controversy" over performing one vs.
two-tailed tests. You really should have some idea going into a study how things
should turn out. Similarly, you should have some inkling of which conditions
should show significant effects.

1.  You're clueless: An omnibus $F$ is significant but you don't know why. In
    this situation, you will perform *post-hoc* tests of means to figure out
    which means are different. If the omnibus $F$ is not significant, *You. Can.
    Not. Perform. Post. Hoc. Tests!*
2.  You know which groups should differ from the others, because you are smart,
    you're an expert, and you gave your experiment lots of thought before you
    collected the data. In this situation you will perform *planned contrasts*
    and you will not conduct an omnibus $F$ test at all.

Here's some new vocabulary: A difference between means or combination of means
is called a *contrast*.

## Post-hoc tests

Post hoc tests identify specific group differences if the ANOVA is significant.
There are lots of options, each with different strengths and weaknesses:

1.  The Bonferroni and Dunn-Šidák procedures
2.  Newman-Keuls test
3.  Scheffé test
4.  Tukey's HSD test
5.  Duncan's multiple range test

We won't discuss them all.

### Bonferroni and Dunn-Šidák

Remember that if we have $J$ groups then, if we want to test all pairs of means,
we will need to perform $M = \binom{J}{2} = J(J-1)/2$ $t$-tests. If we use
something like $\alpha_{PC}=0.05$, where $PC$ indicates "per comparison," then
the probability of making at least one Type I error will be
$\alpha_{FW} = 1 - (1 - \alpha_{PC})^M$, which will be larger than $\alpha_{PC}$
for all $M>1$. We call $\alpha_{FW}$ the *family-wise error rate*. Both the
Dunn-Šidák and Bonferroni procedures "correct" $\alpha_{PC}$ to insure that
$\alpha_{FW}$ is no greater than some desired value (like 0.05).

The Dunn-Šidák correction simply inverts the equation for $\alpha_{FW}$:
\begin{eqnarray*}
\alpha_{FW} & = & 1 - (1 - \alpha_{PC})^M \\
& \rightarrow & (1 - \alpha_{FW}) = (1 - \alpha_{PC})^M \\
& \rightarrow & 1 - \alpha_{PC} = (1 - \alpha_{FW})^{1/M} \\
& \rightarrow & \alpha_{PC} = 1 - (1 - \alpha_{FW})^{1/M}.
\end{eqnarray*}

So if $J = 4$ and our desired $\alpha_{FW} = 0.05$, then $M=6$ and $$
\alpha_{PC} = 1 - (1 - 0.05)^{1/6} = 0.0085.
$$ If we do each of our 6 tests and reject the 6 $H_0s$ only when the $p$-value
is less than $\alpha_{PC} = 0.0085$ then our probability of making one or more
Type I errors will be equal to $\alpha_{FW} = 0.05$

To control the family-wise error rate at level $\alpha$ we can also use the
less-precise but easier Bonferroni correction, which sets the per-comparison
$\alpha_{PC} = \alpha_{FW}/M$, where $M$ is still the number of comparisons to
be made. If again we want $\alpha_{FW} \le 0.05$ and $M=6$, then
$\alpha_{PC} = 0.05/6 = 0.0083$. This value is slightly lower than the
Dunn-Šidák correction, which means that the Bonferroni procedure is more
conservative than the Dunn-Šidák procedure: it will be slightly harder to reject
the $H_0s$ using Bonferroni but the per-comparison $\alpha$-level is easier to
compute.

One important thing to know about pairwise $t$-tests is how the standard error of
the mean difference is computed. Remember for independent t-tests this is $$
s_{\overline{X} - \overline{Y}} = s_P \sqrt{\dfrac{1}{N_X} + \dfrac{1}{N_Y}}
$$ and $$
s_P^2 = \dfrac{(N_X-1) s_X^2 + (N_Y-1) s_Y^2}{N_X + N_Y - 2}.
$$

For the pairwise $t$-tests across more than 2 groups, we will also use a pooled
variance estimate, which we can do because of the homoskedasticity assumption of
the ANOVA model. But now $$
s_P^2 = \dfrac{\sum_{i=1}^J (N_i-1) s_i^2}{N_{\text{Total}} - J},
$$ which is the same thing as $MS_{Within}$. So the standard error of the mean
difference for a pair or means $\overline{X}_i - \overline{X}_j$ *following a
significant omnibus* $F$ is $$
s_{\overline{X}_i - \overline{X}_j} = \sqrt{MS_{Within}} \sqrt{\dfrac{1}{N_i} + \dfrac{1}{N_j}}.
$$

You could, of course, do all these $t$-tests one-by-one by hand. However, `R`
has a tool to perform pairwise t-tests. The command is `pairwise.t.test`.

```{r}
# Compute the per-comparison alpha levels for alpha
# J <- length(unique(health))
M <- choose(J,2) # number of pairs
alpha <- 0.05

# Bonferroni PC error rate
Bonferroni <- alpha / M

# Dunn-Sidak PC error rate
Dunn.Sidak <- 1 - (1 - alpha)^(1/M)
cat("Bonferroni = ",Bonferroni," Dunn-Sidak = ",Dunn.Sidak)

# Perform all M pairwise tests
pairwise.t.test(x=age,g=health,p.adjust.method="none")
```

The `pairwise.t.test` function has a method to adjust $p$-values according to
the desired correction factor. This method is specified by the `p.adjust.method`
option. If you set `p.adjust.method="bonferroni"`, for example, it will multiply
the $t$-test $p$-values by the value of $M$, allowing a comparison between the
adjusted $p$-values and the desired familywise alpha level $\alpha_{FW}$ instead
of the per-comparison alpha level $\alpha_{PC}$. I find this extraordinarily
confusing so I set the $p$-value adjustment to "none." Keep in mind that there
are a large number of possible correction procedures (we have only looked at the
two most common), and so if you find yourself using one we haven't talked about
this option might be useful.

Returning to the output, you can see that we can reject the null hypothesis
$H_0:~\mu_i = \mu_j$ for three comparisons: health=1 vs. health=3 and 4, and
health=2 vs. health=4. There seems to be no significant difference between the
two best levels of health (health=1 and 2), the two poorest levels of health
(health=3 and 4), or between health=2 and health=3.

There are some drawbacks to these two correction procedures. The primary
drawback is that they are extraordinarily conservative. The per-comparison error
rate can be orders of magnitude smaller than the desired familywise error rate,
making it very difficult to reject $H_0$ (and hence greatly reducing power). The
use of the pooled error variance estimate also means that the comparisons may
not be independent, which is an assumption made by both correction methods --
remember that computing $\alpha_{FW} = 1 - (1 - \alpha_{PC})^M$ assumes
independence of the $M$ tests because we're multiplying the probabilities of
*not* making a Type I error on each test together.

For these reasons, alternatives to these correction factors are numerous and
alternatives to pairwise $t$-tests have been developed. Some of these are listed
above. For this notebook we will focus on the popular Tukey's Honestly
Significant Difference test.

### The Honestly Significant Difference

Tukey defined an "honestly significant difference" (HSD) as a single number to
which all mean differences could be compared simultaneously. Among post-hoc
tests, it provides the best balance of power and control of the Type I error
rate.

The HSD test is based on the *Studentized range statistic* $q$, which is defined
as $$
q = \dfrac{\overline{X}_{\max} - \overline{X}_{\min}}{\sqrt{MS_{\text{Within}}/n}},
$$ where the data ($X$) are drawn from a normal distribution,
$\overline{X}_{\max}$ and $\overline{X}_{\min}$ are the largest and smallest
group means, respectively, and $n = N_j, j=1,2,\ldots, J$ is the number of
observations per group (assuming equal group sizes). The statistic $q$ follows a
Studentized range distribution with parameters $J$ (the number of groups) and
$\nu = N_{\text{Total}}-J$ (the degrees of freedom for $MS_{\text{Within}}$).

The distribution of $q$ has root `tukey` in R, but the density function
`dtukey` is not defined because it has no closed form. Only the quantile and
probability functions `qtukey` and `ptukey` are defined. There is a plot of 4
Studentized range distributions (for $k=3,6$ and $n=4,15$) in the lecture notes.
A simulation of the $J=6,~n=15$ case is given in this code chunk. You can play
around with $J$ and $n$ to see how these affect the shape of the distribution.

```{r}
# Simulate the Studentized range distribution for 4 sets of parameters
J <- 6 # number of groups
n <- 15 # group sizes

q <- replicate(10000,{
  # Simulate a dataset
  data <- data.frame(X=rnorm(J*n),group=rep(1:J,each=n))
  # Compute the pooled variance (MS_Within)
  sp2 <- mean(aggregate(data$X,by=list(data$group),var)$x)
  # Compute the values of q
  means <- aggregate(data$X,by=list(data$group),mean)$x
  (max(means)-min(means))/sqrt(sp2/n)
})
hist(q,breaks=seq(min(q)-.1,max(q)+.1,length=50))
```

Now we need to turn the statistic $q$ into an "honestly significant" mean
difference HSD. We do this by first finding the critical value of $q$ that cuts
off our desired $\alpha$ in the upper tail. This is done with the `qtukey`
function. If $J=6$ and $n=15$, this value would be denoted as
$q(1-\alpha,J,\nu = nJ - J)$. (Remember that $\nu$ is the number of degrees of
freedom associated with $MS_{\text{Within}}$.)

```{r}
J <- 6
n <- 15
alpha <- 0.05
qtukey(1-alpha,J,J*n-J)
```

The HSD is the (absolute) mean difference
$\left| \overline{X}_{\max} - \overline{X}_{\min} \right|$ that makes the $q$
statistic greater than the critical value $q(1-\alpha,J,\nu)$. Rearrange
the equation for $q$ to solve for the significant mean difference:
\begin{eqnarray*}
q(1-\alpha,J,\nu)  & = & HSD/\sqrt{MS_{\text{Within}}/n} \\
& \rightarrow & HSD = q(1-\alpha,J,\nu)\sqrt{MS_{\text{Within}}/n}.
\end{eqnarray*}

Continuing with our small example, if $MS_{\text{Within}} = 150$, then $$
HSD = q(1-\alpha,J,\nu) \sqrt{MS_{\text{Within}}/n} = 4.5947 \sqrt{150/15} = 14.97.
$$ If we compute all $\binom{J}{2} = 15$ mean differences, we would need to
compare the absolute values of those differences to the $HSD = 14.97$. All
absolute differences greater than 14.97 would be significant at the
$\alpha = 0.05$ level.

#### Unequal Sample Sizes

An important thing to note is that up to now we have assumed equal group sizes
$n$. If group sizes are unequal we need to figure out what to do with all the
$N_js$. There's a few acceptable strategies for dealing with this, and one is to
use the *harmonic mean* of the group sizes. Another is to compute
$n = 2 N_{\max} N_{\min}/(N_{\max} + N_{\min})$, where $N_{\min}$ and $N_{\max}$
are the sizes of the groups with the largest and smallest means.

What `R` does, by contrast, is called the *Tukey-Kramer method*. Think back to
the Welch's $t$-test, which is the test we used when we couldn't assume
homoskedasticity. For the Welch's test, we estimated the standard error of
$\overline{X}_1- \overline{X}_2$ as
$\sqrt{\dfrac{s_1^2}{N_1} + \dfrac{s_2^2}{N_2}}$. The Tukey-Kramer method uses
the same idea, estimating the standard error of
$\overline{X}_i - \overline{X}_j$ as
$SE_{ij} = \sqrt{\dfrac{1}{2} \left(\dfrac{MS_{\text{Within}}}{N_i} + \dfrac{MS_{\text{Within}}}{N_j}\right)}$.
Thus the HSD will be slightly different for each difference, equal to
$q_{ij}(1 - \alpha,J,\nu) SE_{ij}$.

We'll finish this section by performing Tukey's HSD analysis with the
Tukey-Kramer method, using R's `TukeyHSD` function.

```{r}
tukey.results <- TukeyHSD(age.model)
cat("\n--- Tukey's Honestly Significant Difference (HSD) Results ---\n")
print(tukey.results)
plot(tukey.results)
```

This analysis shows significant differences between health=1 and health=3 and 4,
and health=2 and health=4: the absolute values of these mean differences exceeds
the HSD. `TukeyHSD` also provides a plot, if requested, of each observed value
of $q$ centered within its 95% confidence interval. If the plotted interval does
not include 0 (the dotted vertical line), then the mean difference is
significant at the $\alpha = 0.05$ level.

### 9. Your turn!

Perform a post-hoc analysis of the ANOVA model you fit to your own data. Use the
`pairwise.t.test` and `TukeyHSD` functions. If your omnibus $F$ was not
significant, were any of the mean differences significant? Discuss what will
happen to the Type I error rate if you do post hoc tests after failing to reject
the omnibus hypothesis of all means being equal?

```{r}
# Insert your code here
```

## Planned Contrasts

A contrast is a linear combination of group means used to test a specific
hypothesis about those means. It allows you to focus the test on a theoretically
interesting difference, rather than just every possible pairwise difference. It
is the going into an analysis with an idea of what you want to find that allows
you to bypass the omnibus $F$ test entirely.

We use the notation $\Psi = c_1 \mu_1 + c_2 \mu_2 + \cdots + c_J \mu_J$ to
define a contrast with the goal of testing the null hypothesis $H_0:~\Psi = 0$.
The contrast coefficients $c_1, c_2, \ldots, c_J$ must sum to 0. The pairwise
$t$-tests we did for post hoc tests are defined by one $c_i=1$, another $c_j=-1$
and all the other $c_ks=0$. For example, we showed that the mean age for
health=1 and health=3 were significantly different. This contrast is defined by
the coefficient vector $(1,0,-1,0)$, or
$\Psi = (1) \mu_1 + (0) \mu_2 + (-1) \mu_3 + (0) \mu_4 = \mu_1 - \mu_3$.

My hypothesis is that people who report better levels of reported health
(health=1 and 2) are younger than those who report poorer levels of reported
health (health=3 and 4), because aging generally has negative impacts on overall
health. I planned to test this hypothesis in advance of collecting the data, and
the contrast that describes this pattern of results is defined by the
coefficient vector $c(1/2,1/2,-1/2,-1/2)$. Note that you have some flexibility
in selecting the values of the coefficients. They should appropriately weight
the means to reflect the hypotheses of interest. It is common practice to select
them so that the absolute value sums to 2, consistent with $\Psi$ representing a
difference between two averaged groups.

The coefficient vector $c(1/2,1/2,-1/2,-1/2)$ defines the contrast
$\Psi = (1/2)\mu_1 + (1/2)\mu_2 - (1/2) \mu_3 - (1/2)\mu_4 = \dfrac{\mu_1 + \mu_2}{2} - \dfrac{\mu_3 + \mu_4}{2}$:
if $\Psi = 0$ then the mean of the two best health groups is equal to the mean
of the two poorest health groups.

But how do we test $H_0:~\Psi = 0$? Turns out that's pretty simple. First we
compute the observed value of the contrast, which is $$
\hat{\Psi} = \sum_{j=1}^J c_j \overline{X}_j.
$$ Then we need to compute the sum of squares for the contrast $\Psi$, which
should equal $MS_{\text{Within}}$ approximately if $\Psi=0$. The sum of squares
for $\Psi$ is given by $SS_{\Psi} = \hat{\Psi}/w$, where
$w = \sum_{j=1}^J c_j^2/N_j$. This sum of squares has only 1 degree of freedom,
so $MS_{\Psi} = SS_{\Psi}$.

Finally, we compute the $F$-ratio for the contrast as
$MS_{\Psi}/MS_{\text{Within}}$, which follows an $F$ distribution with 1
numerator and $N_{\text{Total}}-J$ degrees of freedom.

Here's how we do this in R. There's a useful package, `multcomp`, that makes
this easy. First we'll do the computations by hand, then we'll use the package.

```{r}
# Define the contrasts of interest
contrast.mx <- rbind(
  "Good health vs Poor health" = c(.5, .5, -.5, -.5),  
  "Best health vs Rest" = c(1, -1/3, -1/3, -1/3)
)
# Group means and group sizes were computed earlier
# group.means <- aggregate(age,by=list(health),mean)$x
# N.groups <- as.numeric(table(health))

# Pull out MS_Within from the ANOVA summary 
MS.Within <- summary(age.model)[[1]][["Mean Sq"]][2]

# Compute the p-value of the first contrast by hand
Psi.hat <- contrast.mx %*% group.means
w <- rowSums(contrast.mx^2 / rbind(N.groups,N.groups))
SS.Psi <- (Psi.hat)^2 / w
MS.Psi <- SS.Psi # 1 df
F.value <- MS.Psi / MS.Within
p.value <- 1 - pf(F.value,1,sum(N.groups)-length(N.groups))
data.frame(Psi.hat,MS.Psi,F.value,p.value)
```

Now use the multcomp package to do the contrasts.

```{r}
library(multcomp)
# Use the glht and mcp functions to compute contrasts
age.glht <- glht(age.model,linfct=mcp(health=contrast.mx))
summary(age.glht,test=adjusted("none"))
```

The output from the glht function looks a little different from what we computed
by hand. There are several reasons for this, but be reassured that the two
analyses are equivalent.

First, the glht function computes t-values rather than F-values. This is okay,
because we know that a squared $t$ with $J$ degrees of freedom is equal to an
$F$ with 1 and $J$ degrees of freedom. You can verify that the $t$-statistics
output by glht are equal to the square roots of the $F$-values we computed by
hand.

```{r}
# Verify that t^2 = F
cat("From the glht function:\n")
summary(age.glht)$test$tstat^2
cat("\nFrom our by-hand calculations:\n")
F.value
```

Second, the glht function provides standard errors rather than mean squares.
Again, this is okay because we can compute mean squares from standard errors.
The standard error of the $t$ statistic for a contrast $\Psi$ is computed as
$SE_{\Psi} = \sqrt{MS_{\text{Within}} w}$. Thus we can compute the mean square
for the contrast as $MS_{\Psi} = SE_{\Psi}^2 / w$. You can verify that the mean
squares we computed by hand are equal to those computed from the standard errors
output by glht.

```{r}
# Verify that MS.Within = SE.Psi^2 / w and that SE.Psi = sqrt(MS.Within * w)
# Standard errors from the glht output
SE.Psi.glht <-summary(age.glht)$test$sigma
SE.Psi.glht

# Compute the standard errors from the MS.Within and the w we computed by hand
sqrt(MS.Within * w)

# Compute MS.Within from the standard errors in the glht output
SE.Psi.glht^2 / w
MS.Within
```

Finally, note that the p-values output by glht are the same as those we computed
by hand.

```{r}
# Verify that p-values are almost equal
cat("From the glht function:\n")
summary(age.glht,test=adjusted("none"))$test$pvalues
cat("\nFrom our by-hand calculations (times 2 for two-tailed):\n")
p.value
```

OK, but wait! Didn't we just two two $t$-tests (and two $F$-tests)? Aren't we
supposed to adjust our alpha level if we do multiple tests? Yes, yes we are. The
`glht` function makes this adjustment via that `test=adjusted()` option in the
`summary` function. By saying "none" for this we forced `summary` not to adjust
for multiple comparisons. By default `ghlt` uses the something called a
"single-step method" (similar to the Bonferroni correction) to inflate the
$p$-values and penalize us appropriately for multiple tests.

```{r}
summary(age.glht)
```

Compare these adjusted $p$-values to those we computed using the unadjusted
method. Note that the $p$-values are larger now, approximately twice those of
the unadjusted values, reflecting the penalty for multiple tests.

### 10. Your turn!

Perform planned contrasts on your own data. Define at least two contrasts that
reflect interesting hypotheses about your data. Use either the by-hand method or
the `glht` function from the `multcomp` package. Explain your results.

```{r}
# Insert your code here
```

## Orthogonal Contrasts

When you perform multiple planned contrasts, it is important to consider whether
the contrasts are *orthogonal* or not. Orthogonal contrasts are statistically
independent of one another, meaning that the outcome of one contrast does not
influence the outcome of another. This independence is important because it
allows for clearer interpretation of results and helps maintain the overall Type
I error rate when multiple contrasts are tested.

Two contrasts defined by coefficient vectors $c^{(1)}$ and $c^{(2)}$ are
orthogonal if $$\sum_{j=1}^J N_j c_j^{(1)} c_j^{(2)} = 0.$$ If the group sizes
are equal, this reduces to the simpler condition
$$\sum_{j=1}^J c_j^{(1)} c_j^{(2)} = 0.$$

For example, consider the two contrasts we defined earlier for the health
variable: 1. Good health vs Poor health: $c^{(1)} = (0.5, 0.5, -0.5, -0.5)$ 2.
Best health vs Rest: $c^{(2)} = (1, -1/3, -1/3, -1/3)$ To check if these
contrasts are orthogonal, we compute the sum of the products of their
coefficients: $$
(0.5)(1) + (0.5)(-1/3) + (-0.5)(-1/3) + (-0.5)(-1/3) = 0.5 - 0.1667 + 0.1667 + 0.1667 = 0.6667 \neq 0.$$
(Weighting by sample sizes also results in a sum that is not equal to 0.)
Because the sum is not zero, these contrasts are not orthogonal.

We talked about partitioning total variance in ANOVA. When contrasts are
orthogonal, the sums of squares associated with each contrast adds up to the
total sum of squares between groups. This means that each contrast explains a
unique portion of the variance in the data, allowing for a clearer understanding
of how different factors contribute to group differences.

When contrasts are not orthogonal, they do not cleanly partition the variance.
Their variances overlap and the sum of squares for the contrasts will not add up
to the total sum of squares between groups, making it harder to discern the
unique contribution of each contrast.

When planning contrasts, it is often beneficial to design them to be orthogonal,
especially when multiple contrasts are being tested. This helps ensure that each
contrast provides distinct information about the group differences and maintains
the integrity of statistical inferences. However, not all research questions can
be addressed with orthogonal contrasts; sometimes non-orthogonal contrasts are
necessary to test specific hypotheses.

For $J$ groups there are at most $J-1$ orthogonal contrasts. There are different
kinds of orthogonal contrasts, including polynomial contrasts (which test for
linear, quadratic, cubic, etc. trends across ordered groups) and Helmert
contrasts (which compare each group mean to the mean of subsequent groups). You
can also create your own custom contrasts designed to test specific hypotheses
while maintaining orthogonality.

```{r}
# Define orthogonal contrasts for 4 groups
poly.contrasts <- contr.poly(4)
helmert.contrasts <- contr.helmert(4)
print("Polynomial Contrasts:")
print(poly.contrasts)
print("Helmert Contrasts:")
print(helmert.contrasts)
```

The `contrasts` function in `R` can be used to incorporate contrasts into an
ANOVA. The `split` option in the `summary` function tells `R` to break down the
sum of squares into components associated with each contrast.

```{r}
# Fit ANOVA model with polynomial contrasts
contrasts(health) <- poly.contrasts
age.poly.model <- aov(age ~ health)
summary(age.poly.model,
        split=list(health=list("Linear"=1,
                              "Quadratic"=2,
                              "Cubic"=3)))
```

Setting $\alpha_{PC} = 0.05/3 = 0.0167$ to control the family-wise error rate at
$\alpha_{FW} = 0.05$ for the three contrasts, we see that only the linear
contrast is significant. This analysis suggest that the variance in age across
health levels is primarily explained by a linear trend, with no significant
quadratic or cubic components.

Note that the `split` option only works when the contrasts are orthogonal. If
you try to use it with non-orthogonal contrasts or with unequal group sizes the
values that `summary` returns will not be the same as those that you compute by
hand. If you want to use non-orthogonal contrasts or have unequal group sizes,
you should use the `glht` function from the `multcomp` package as we did
earlier.

### 11. Challenge: Your turn!

Define at least two orthogonal contrasts for your own data. You can use
polynomial, Helmert, or your own orthogonal contrasts. Perform the analysis
using either the `split` option in the `summary` function or the `glht` function
from the `multcomp` package. Explain your results.

```{r}
# Insert your code here
```

## Summary

Post hoc and planned contrasts allow you to test specific hypotheses about group
means. Post hoc contrasts require a significant omnibus F test; planned
contrasts do not require this.

Bonferroni and Dunn-Šidák corrections are simple ways to control the family-wise
error rate $\alpha_{FW}$ when performing multiple post hoc tests, but they are
conservative and reduce power. Tukey's HSD test provides a better balance of
power and Type I error control for post hoc tests and does not require
adjustment to the per-comparison alpha level $\alpha_{PC}$.

Planned contrasts, although they do not require a significant omnibus F test,
will require adjustments for multiple comparisons. Bonferroni or Dunn-Šidák
corrections can be used here as well, but you might consider using the `glht`
function from the `multcomp` package to do the heavy lifting.

The use of orthogonal contrasts and post hoc pairwise t-tests share the
characteristic that neither approach is based on a theoretical understanding of
the data. Orthogonal contrasts are statistically independent, but they may not
reflect the hypotheses of interest. Pairwise $t$-tests examine all possible mean
differences, but many of these may not be theoretically interesting. When
possible, it is best to define planned contrasts that reflect your theoretical
understanding of the data, and remember to adjust your $\alpha_{PC}$ level to
maintain a low $\alpha_{FW}$ error rate.

# Factorial ANOVA

To this point we have been doing "one-way" ANOVAs with a single categorical
independent variable. Now we'll expand that to two. When we have observations in
each combination of levels of two categorical independent variables, we call
this a "two-way" or "factorial" ANOVA. Factorial ANOVAs allow us to examine not
only the *main effects* of each independent variable on the dependent variable,
but also whether there is an *interaction effect* between the two independent
variables. We could have more than two independent variables, of course, but for
now we'll stick with two for simplicity. The principles we discuss here will
generalize to more than two independent variables.

Let's load up the data again and start from scratch.

```{r}
data <- read.csv("GSS_1980.csv",header=TRUE)
# Pull out variables of interest
happy <- data$HAPPY
age <- data$AGE
health <- data$HEALTH
```

For the functions we'll use later, it's important to have equal sample sizes, so
we'll first collapse health=3 and 4 into a single category ("not good") and then
pull out 25 observations from each combination of health and happy.

```{r}
# Collapse health=3 and 4 into a single category
health.sub <- health
health.sub[health==4] <- 3

# Change health and happy to factors
health.sub <- factor(health.sub)
happy.sub <- factor(happy)
```

Subset the data to ensure equal sample sizes in each combination of health and
happy.

```{r}
set.seed(1234) # for reproducibility
# Pull 25 observations from each combination of health and happy and make a new dataset from them
data.sub <- data.frame()
# For each level of health
for (i in levels(health.sub)) {
  # and each level of happy
  for (j in levels(happy.sub)) {
    # extract all observations in the cell
    combo.obs <- which(health.sub == i & happy.sub == j)
    # and choose 25 of them at random.
    if (length(combo.obs) > 25) obs <- sample(combo.obs, 25,replace=FALSE)
    # Add them to the new data frame
    data.sub <- rbind(data.sub,
                      data.frame(health=rep(i,length(obs)),
                                 happy=rep(j,length(obs)),
                                 age=age[obs]))
  }
}
# Pull the variables out of the data frame so we don't have to type $ signs
age.sub <- data.sub$age
health.sub <- data.sub$health
happy.sub <- data.sub$happy
```

Take a quick look at the subsetted data.

```{r}
# Make a box and whisker plot
Desc(age.sub ~ health.sub + happy.sub, 
     main="Age by Health and Happiness",
     xlab="Health and Happiness Levels",
     ylab="Age")
```

The difference between a one-way and a two-way ANOVA is in how the sums of
squares are partitioned. In a one-way ANOVA we partition the total sum of
squares into between groups $SS_{\text{Between}}$ and within groups
$SS_{\text{Within}}$ sums of squares. In a two-way ANOVA we partition the total
sum of squares into three between groups sums of squares (one for each
independent variable and one for the interaction between them) and the within
groups sum of squares.

To partition the variance in this kind of design, consider first the contingency
that counts the number of observations in each combination of the levels of each
independent variable.

```{r}
table("J"=health.sub,"K"=happy.sub)
```

I renamed the two independent variables $J$ and $K$ so that we can understand
the notation a little better. I'm also going to let $J$ and $K$ represent the
number of levels of each factor. It should be clear which I mean whenever I use
it (and if it isn't please let me know.)

Each frequency (25) of the pairs of levels of $J$ and $K$ in the table above
represents a "cell" of the experiment. In this case there are $J=3$ levels of
health and $K=3$ levels of happy, for a total of $J \times K = 9$ cells. Each
cell has $N_{jk}$ observations, where $j=1,2,\ldots,J$ and $k=1,2,\ldots,K$. The
total number of observations across all cells is
$N_{\text{Total}} = \sum_{j=1}^J \sum_{k=1}^K N_{jk}$. An individual measurement
in the $jk^{th}$ cell is denoted as $X_{ijk}$, where $i=1,2,\ldots,N_{jk}$
indexes the individual observations within that cell.

Now things will get a little strange but stick with me. We need to define
several means to help us partition the variance. The grand mean of all
observations across all cells is
$\overline{\overline{X}} = \dfrac{1}{N_{\text{Total}}} \sum_{j=1}^J \sum_{k=1}^K \sum_{i=1}^{N_{jk}} X_{ijk}$.
Note the addition of a new summation sign that corresponds to the sum over the
second independent variable $K$.

We need to compute *marginal* sample sizes and means for each level of each
factor. The notation will use a dot to indicate the factor over which the sample
sizes and means are being taken. So, for example, the number of observations in
level $j$ of variable $J$ will be written as $N_{j \cdot}$. The means of the
observations at level $j$ of factor $J$ is
$\overline{X}_{j\cdot} = \dfrac{1}{N_{j \cdot}} \sum_{k=1}^K \sum_{i=1}^{N_{jk}} X_{ijk}$.
The mean of the observations at level $k$ of factor $K$ is similarly
$\overline{X}_{\cdot k} = \dfrac{1}{N_{\dot k}} \sum_{j=1}^J \sum_{i=1}^{N_{jk}} X_{ijk}$.

Finally, we need the cell means. The mean of the observations in the $jk^{th}$
cell is $\overline{X}_{jk} = \dfrac{1}{N_{jk}} \sum_{i=1}^{N_{jk}} X_{ijk}$.

Consider again the total sum of squares $SS_{\text{Total}}$. We will compute
this the same way as before: $$
SS_{\text{Total}} = \sum_{k=1}^K \sum_{j=1}^J \sum_{i=1}^{N_{jk}} \left(X_{ijk} - \overline{\overline{X}} \right)^2
$$ Note that this is just the numerator of $s^2_X$ without regard to any group
structure.

```{r}
# Compute the grand mean and total variance
N.Total <- length(age.sub)
grand.mean <- mean(age.sub)
grand.var <- var(age.sub)

# Compute the total sum of squares
SS.Total <- sum((age.sub - grand.mean)^2)

cat("Grand Mean:", grand.mean, "\n")
cat("Total Sum of Squares:", SS.Total, "\n")

# Notice that SS.Total = (N.Total - 1) * grand.var
cat("Verification:", (N.Total - 1) * grand.var, "\n")
```

In one-way ANOVA we talked about $SS_{\text{Between}}$ and $SS_{\text{Within}}$.
In two-way ANOVA the $SS_{\text{Between}}$ will be replaced with
$SS_{\text{Cells}}$, which is computed in the same way: $$
SS_{\text{Cells}} = \sum_{k=1}^K \sum_{j=1}^J N_{jk} \left( \overline{X}_{jk} - \overline{\overline{X}} \right)^2.
$$

But now, we will partition the $SS_{\text{Cells}}$ into three components: the
sum of squares $SS_J$ associated with factor $J$, the sum of squares $SS_K$
associated with factor $K$, and the sum of squares $SS_{JK}$ associated with the
interaction between factors $J$ and $K$. These are computed as follows: $$
SS_J = \sum_{j=1}^J N_{j \cdot} \left( \overline{X}_{j \cdot} - \overline{\overline{X}} \right)^2,
$$
$$SS_K = \sum_{k=1}^K N_{\cdot k} \left( \overline{X}_{\cdot k} - \overline{\overline{X}} \right)^2,
$$ and then, because the interaction term is what's left over after accounting
for the main effects of $J$ and $K$,
$$SS_{J \times K} = SS_{\text{Cells}} - (SS_J + SS_K).
$$

The only thing we have to compute now is the within groups sum of squares, which
can be computed by pooling the cell variances $s^2_{ij}$ the same way as before.
However, knowing that
$SS_{\text{Total}} + SS_{\text{Cells}} + SS_{\text{Within}}$, $$
SS_{\text{Within}} = SS_{\text{Total}} - SS_{\text{Cells}},
$$ which is a bit easier.

Treat the age variable $X$ as a function of health $J$ and happiness $K$. We
might expect to see age increase with increases in health (remember that
health=1 is excellent health and health=4 is poor health). We might also expect
to see differences in age with reported happiness (happy people might tend to be
older because they are more stable, or unhappy people might tend to be older
because of poor health or having lost friends and family). There might also be
an *interaction* between health and happiness, such that the age difference
between happy and less happy people might be less for unhealthy than for healthy
people.

Let's fit the ANOVA model and look for these sums of squares in the output. For
the `aov` function, we specify two independent variables by expanding the model
formula from `age.sub ~ health.sub` to `age.sub ~ health.sub + happy.sub + health.sub:happy.sub`, where
happy is the second independent variable and `health.sub:happy.sub` indicates an
interaction between health and happy. The `+` signs indicate that we are
including main effects for both health and happy, while the `:` indicates that
we are including an interaction effect between health and happy.  An equivalent formula is `age.sub~health.sub*happy.sub`, which includes main effects and the interaction explicitly.

```{r}
# Fit the two-way ANOVA model
age.factorial.model <- aov(age.sub ~ 
                    happy.sub + health.sub + 
                    health.sub:happy.sub)
summary(age.factorial.model)
```

You can validate the degrees of freedom and sums of squares computed by `R` by
computing them by hand. This chunk of code computes the degrees of freedom and
sums of squares for the main effects, interaction, and error
($SS_{\text{Within}}$). For each main effect, if there are $J$ and $K$ levels of
factors $J$ and $K$, respectively, then the degrees of freedom for each main
effect is $df_J = J-1$ and $df_K = K-1$. The degrees of freedom for the
interaction effect is $df_{J \times K} = (J-1)(K-1)$. The degrees of freedom for
the within groups variance is $df_{\text{Within}} = N_{\text{Total}} - JK$; 
think about this one as you did for the one-way ANOVA where you subtracted off
the number of levels, but now you're subtracting off the number of cells. The
sum of the main effects and interaction degrees of freedom should equal the
total degrees of freedom between groups:
$df_J + df_K + df_{J \times K} = JK - 1$.

```{r}
# Compute cell means
cell.means <- aggregate(age.sub, by=list(health.sub, happy.sub), mean)
colnames(cell.means) <- c("health", "happy", "mean")

# Compute marginal means
marginal.means.health <- aggregate(age.sub, by=list(health.sub), mean)
colnames(marginal.means.health) <- c("health", "mean")
marginal.means.happy <- aggregate(age.sub, by=list(happy.sub), mean)
colnames(marginal.means.happy) <- c("happy", "mean")

# Compute sample sizes
N.jk <- as.matrix(table(health.sub, happy.sub))
N.j.dot <- rowSums(N.jk)
N.dot.k <- colSums(N.jk)

# Compute the degrees of freedom
df.Health <- length(N.j.dot)-1 # J-1
df.Happy <- length(N.dot.k)-1 # K-1
df.HealthHappy <- df.Health*df.Happy # (J-1)(K-1)
df.Within <- N.Total - length(N.j.dot)*length(N.dot.k) # N.Total - JK

# Compute SS_Cells
SS.Cells <- sum(N.jk * (matrix(cell.means$mean, 
                nrow=3, ncol=3) - grand.mean)^2)
# Compute SS_Health
SS.Health <- sum(N.j.dot * (marginal.means.health$mean - grand.mean)^2)
# Compute SS_Happy
SS.Happy <- sum(N.dot.k * (marginal.means.happy$mean - grand.mean)^2)
# Compute SS_Interaction
SS.Interaction <- SS.Cells - (SS.Health + SS.Happy)
# Compute SS_Within
SS.Within <- SS.Total - SS.Cells

# Display results so far
factorial.anova <- data.frame("Source" = c("Health","Happy","Health:Happy","Within"),
          "Df"=c(df.Health,df.Happy,df.HealthHappy,df.Within),
          "SS" = c(SS.Health,SS.Happy,SS.Interaction,SS.Within))
factorial.anova
```

To complete the ANOVA table we need to compute the mean squares and F-ratios for
each source of variance. The mean squares are computed as usual:
$MS_{\text{Source}} = SS_{\text{Source}}/df$. The F-ratios are also computed as
usual: $F = MS_{\text{Source}}/MS_{\text{Within}}$.

```{r}
# Compute the mean squares
MS.Health <- SS.Health / df.Health
MS.Happy <- SS.Happy / df.Happy
MS.Interaction <- SS.Interaction / df.HealthHappy
MS.Within <- SS.Within / df.Within
factorial.anova$MS <- c(MS.Health, MS.Happy, MS.Interaction, MS.Within)

# Compute the F ratios
F.Health <- MS.Health / MS.Within
F.Happy <- MS.Happy / MS.Within
F.Interaction <- MS.Interaction / MS.Within
factorial.anova$F <- c(F.Health, F.Happy, F.Interaction,NA)

# Compute the F p-values; note the degrees of freedom
p.Health <- 1 - pf(F.Health, df.Health, df.Within)
p.Happy <- 1 - pf(F.Happy, df.Happy, df.Within)
p.Interaction <- 1 - pf(F.Interaction, df.HealthHappy, df.Within)
factorial.anova$p.value <- c(p.Health, p.Happy, p.Interaction,NA) 
factorial.anova
```

Check now to see that the `factorial.anova` data frame gives the same values as
the `summary` output.

### Unequal Sample Sizes

If the experiment is not balanced (equal sample sizes in each cell), because the
`aov` function returns what is called "Type I" sums of squares, they will not be
equal to the sums of squares you compute by hand. Furthermore, the sums of
squares will change depending on the order of the variables in the formula
passed to `aov`. This is because the sums of squares for each source can no
longer be perfectly partitioned: there will be some influence of factor $J$ in
the main effect of $K$ and vice versa.

Go back to the full GSS_1980 dataset.  Fit the factorial ANOVA model in two ways, with health first and then with happy first.
```{r}
# Fit the two-way ANOVA model with health entered first
summary(aov(age~health+happy+health:happy))

# Now put happy first
summary(aov(age~happy+health + health:happy))
```

If you compute the sums of squares by hand you will see that the sums of squares
for health equals the sums of squares computed by `aov` when health is first.
The sums of squares for happy equals the sums of squares computed by `aov` when
happy is first.

A detailed explanation for why this happens is beyond the scope of this notebook
(but check
[this](https://bookdown.org/ndphillips/YaRrr/type-i-type-ii-and-type-iii-anovas.html)
link). Briefly, Type I sums of squares are computed by fitting different models
sequentially, starting with the simplest and adding factors one by one. Type III
sums of squares are computed by fitting all factors first and then removing them
one at a time. There are also Type II and IV sums of squares. Each type of sums
of squares has its uses, but Types II and III are generally preferred for
unbalanced designs because the results are independent of the order in which the
factors are specified in the model formula.

For unbalanced designs you can use the `Anova` function from the `car` package,
which computes Type II and III sums of squares that are independent of the order
in which the independent variables are specified in the model formula. I've
always used Type III sums of squares, but the reason for this is [probably based
mostly in
tradition](https://seriousstats.wordpress.com/2020/05/13/type-ii-and-type-iii-sums-of-squares-what-should-i-choose/)
and not any strong theoretical foundation.

```{r}
# Fit the two-way ANOVA model to the whole dataset using lm() and then Anova() from the car package
age.Anova <- lm(age~health+happy+happy:health)
Anova(age.Anova,type="III")
```

Only the main effect of `happy` is significant at an $\alpha=0.05$ level.  But wait...

### How Many Tests?

You probably wondered about doing three $F$-tests and the Type I error inflation problem we talked about earlier in the context of pairwise $t$-tests.  Good for you!

Yes, [you should correct your alpha level to accommodate 3 tests](https://link.springer.com/article/10.3758/s13423-015-0913-5).  Be aware that people will argue with you about this.  The conservative and statistically "pure" approach would be to reduce the per-comparison alpha level according to something like the Bonferroni or Dunn-Šidák correction.  Common practice, however, is to argue that the ANOVA model you just fit is a single test, the sums of squares are independent from each other (if the group sizes are equal), and performing a correction is "too conservative."  

If you use a Bonferroni correction for the alpha-level in the ANOVA you just performed, $\alpha_{PC} = 0.05/3 = 0.0167.$  In this case, there are no significant main effects of or interactions between health and happy on respondent age in the full data set.  In the subsetted (balanced) dataset the main effect of happy is still significant after the correction.

### 12. Your turn!

Choose two independent variables and one dependent variable and plot the means
of the dependent variable as a function of the values of the two independent
variables.

1.  What does the mean pattern look like? Are there any main effects? Is there
    an interaction?
2.  Fit an ANOVA model predicting the dependent variable from the two
    independent variables and their interaction. Use `aov` (if there are roughly
    equal group sizes) or `Anova` if the groups are unequal. Describe your
    results and provide an interpretation. Are the significant (and
    insignificant) effects consistent with what you saw in your plots?

```{r}
# Insert your code here
```

# Repeated Measures ANOVA

Repeated measures ANOVA is used when the same subjects are measured under
different conditions or at different time points. This design accounts for the
correlation between measurements taken from the same subject, which can increase
statistical power.

For this section we'll use some (simulated) recognition memory data. Five
participants are given 3 lists of 40 words each to memorize. Depending on the
list, they are limited to 500ms, 1000ms, or 1500ms per word. At test, the words
just studied are mixed with 40 new words and shown to the participants one at a
time. They are told to respond "yes" if the word was studied and "no" if the
word is new. The number of correct responses is recorded.

```{r}
memory <- c(43, 52, 62, 47, 65, 75, 62, 71, 79, 55, 67, 71, 36, 46, 57)
participant <- as.factor(rep(1:5,each=3))
times <- as.factor(rep(c(500,1000,1500),times=5))
```

For repeated measures designs, the participant factor is treated like an
independent variable, but we say that the time factor is "nested" within it. (Variables like "participant" are called *random effects*, and models that include both fixed factors (like times) and random effects (like participant) are called mixed-effects models.)  The
way that the sums of squares are partitioned is determined by the formula passed
to `aov` (or `Anova`). The big difference between a one-way ANOVA and a
(one-way) repeated measures ANOVA is that for the repeated measures ANOVA we
have to tell `aov` what the error term ($MS_{\text{Error}}$) is. It isn't
$MS_{\text{Within}}$ any more.

```{r}
summary(aov(memory~times+Error(participant/times)))
```

The output shows that the main effect of time is significant at an $\alpha=0.05$
level.

### 13. Challenge/Extra: Your turn!

If you have a repeated measure in your dataset, perform an ANOVA to determine if
that treatment has a significant effect on your dependent measure. Describe your
results.

```{r}
# Insert your code here
```

# 14. Conclusions

Think about what you've done in this notebook.

a. What was the most important thing you learned?
b. What was the hardest concept to understand?
c. What surprised you the most?

> Insert your responses here

```{r}
# Insert your code here
```
